---
title: "Data Wrangling - Tidyr"
author: "Novi s adi"
date: "10/5/2021"
output: html_document
---

# Data wrangling with `tidyr`, which is part of the tidyverse. We are going to tidy some data!

```{r}
## wide format
gap_wide <- readr::read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder_wide.csv')
```
# 1. Not satisfacgtory
```{r}
library (tidyverse)
gap_long <- gap_wide %>% gather(key = obstype_year,
         value = obs_values)
```

# 2. Too much argument
```{r}
gap_long <- gap_wide %>% 
  gather(key   = obstype_year,
         value = obs_values,
         dplyr::starts_with('pop'),
         dplyr::starts_with('lifeExp'),
         dplyr::starts_with('gdpPercap'))  #here i'm listing all the columns to use in gather
```

# 3. Most effective one
```{r}
gap_long <- gap_wide %>% 
  gather(key   = obstype_year,
         value = obs_values,
         -continent, -country)
```

# IMPORTANT !!!
```{r}
gap_long <- gap_wide %>% 
  gather(key   = obstype_year,
         value = obs_values,
         -continent, -country) %>%
  separate(obstype_year,
           into = c('obs_type','year'),
           sep = "_",
           convert = TRUE); gap_long #this ensures that the year column is an integer rather than a character
```

# 7.5 Plot long format data
The long format is the preferred format for plotting with ggplot2. Let’s look at an example by plotting just Canada’s life expectancy.

```{r}
canada_df <- gap_long %>%
  filter(obs_type == "lifeExp",
         country == "Canada")
```

```{r}
ggplot(canada_df, aes(x = year, y = obs_values)) +
  geom_line()
```

We can also look at all countries in the Americas:

```{r}
life_df <- gap_long %>% filter (obs_type == "lifeExp",continent == "Americas")
```

```{r}
ggplot(life_df, aes(x = year, y = obs_values, color = country)) +
  geom_line()
```

# 7.5 Exercise

Using gap_long, calculate and plot the the mean life expectancy for each continent over time from 1982 to 2007. Give your plot a title and assign x and y labels. Hint: do this in two steps. First, do the logic and calculations using dplyr::group_by() and dplyr::summarize(). Second, plot using ggplot()

# Alternative 1
```{r}
continents <- gap_long %>% group_by (continent, year) %>% filter (obs_type == "lifeExp", year >= 1982 & year <= 2007) %>% summarize(mean_le = mean(obs_values)) %>% 
  ungroup(); continents
```

# Alternativ 2
```{r}
continents <- gap_long %>% group_by (continent, year) %>% filter (obs_type == "lifeExp", year >= 1982 & year <= 2007) %>% summarize(mean_le = mean(obs_values)) %>% arrange (desc (mean_le)) %>%
  ungroup(); continents
```
```{r}
# plotting opsi 1
ggplot(data = continents, aes(x = year, y = mean_le, color = continent)) + 
  geom_line() +
  labs(title = "Mean life expectancy",
       x = "Year",
       y = "Age (years)") 
```

```{r}
# plotting opsi 2
ggplot(data = continents, aes(x = year, y = mean_le, color = continent)) + 
  geom_line() +
  labs(title = "Mean life expectancy",
       x = "Year",
       y = "Age (years)",
       color = "Continent") +
  theme_classic() +
  scale_fill_brewer(palette = "Blues")  
```

# 7.6 spread()

```{r}
gap_normal <- gap_long %>% 
  spread(obs_type, obs_values); gap_normal
```

# 7.6 Exercise

1. Convert gap_long all the way back to gap_wide. Hint: Do this in 2 steps. First, create appropriate labels for all our new variables (variable_year combinations) with the opposite of separate: tidyr::unite(). Second, spread() that variable_year column into wider format.

2. Knit the R Markdown file and sync to Github (pull, stage, commit, push)

7.6.1 Answer

#steps 
```{r}
gap_wide_new <- gap_long %>% 
  # first unite obs_type and year into a new column called var_names. Separate by _
  unite(col = var_names, obs_type, year, sep = "_") %>% 
  # then spread var_names out by key-value pair.
  spread(key = var_names, value = obs_values); gap_wide_new
```

# 7.7.1 complete()

```{r}
kelpdf <- data.frame(
  Year = c(1999, 2000, 2004, 1999, 2004),
  Taxon = c("Saccharina", "Saccharina", "Saccharina", "Agarum", "Agarum"),
  Abundance = c(4,5,2,1,8)
) ; kelpdf
```

Jarrett points out that Agarum is not listed for the year 2000. Does this mean it wasn’t observed (Abundance = 0) or that it wasn’t recorded (Abundance = NA)? Only the person who recorded the data knows, but let’s assume that the this means the Abundance was 0 for that year.

We can use the complete() function to make our dataset more complete.

```{r}
kelpdf %>% 
  complete(Year, Taxon)
```

This gives us an NA for Agarum in 2000, but we want it to be a 0 instead. We can use the fill argument to assign the fill value.

```{r}
kelpdf %>% complete(Year, Taxon, fill = list(Abundance = 0))
```
Now we have what we want. Let’s assume that all years between 1999 and 2004 that aren’t listed should actually be assigned a value of 0. We can use the full_seq() function from tidyr to fill out our dataset with all years 1999-2004 and assign Abundance values of 0 to those years & species for which there was no observation.

```{r}
kelpdf %>% complete(Year = full_seq(Year, period = 1), Taxon, fill = list(Abundance = 0))
```

