---
title: "data wrangling using tidyverse and dplyr"
author: "Novi s adi"
date: "10/2/2021"
output: html_document
---


```{r}
library(tidyverse)     ## install.packages("tidyverse")
```

```{r}
## read gapminder csv. Note the readr:: prefix identifies which package it's in
gapminder <- readr::read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv') 
```

```{r}
## explore the gapminder dataset
gapminder # this is super long! Let's inspect in different ways
```

```{r}
head(gapminder) # shows first 6
tail(gapminder) # shows last 6
```

```{r}
head(gapminder, 10) # shows first 10
tail(gapminder, 12) # shows last 12
```

```{r}
str(gapminder) # ?str - displays the structure of an object
```

```{r}
names(gapminder)
dim(gapminder)    # ?dim dimension
ncol(gapminder)   # ?ncol number of columns
nrow(gapminder)   # ?nrow number of rows
```

```{r}
summary(gapminder)
```

```{r}
library(skimr) # install.packages('skimr')
```

```{r}
skim(gapminder)
```

```{r}
gapminder$lifeExp # very long! hard to make sense of...
```

```{r}
head(gapminder$lifeExp) # can do the same tests we tried before
```

```{r}
str(gapminder$lifeExp) # it is a single numeric vector
```

```{r}
summary(gapminder$lifeExp) # same information, formatted slightly differently
```

# 6.4 dplyr basics

## filter() is a function in dplyr that takes logical expressions and returns the rows for which all are TRUE.

```{r}
filter(gapminder, lifeExp < 29)
```

```{r}
filter(gapminder, country == "Mexico")
```

```{r}
filter(gapminder, country %in% c("Mexico", "Peru"))
```

```{r}
filter(gapminder, country == "Mexico", year == 2002)
```

# 6.6 Your turn

## What was the average life expectency in Brazil between 1987 and 2007? 
## Hint: do this in 2 steps by assigning a variable and then using the mean() function.
## Then, sync to Github.com (pull, stage, commit, push).

```{r}
b <- filter (gapminder, country == "Brazil", year > 1986)
```

```{r}
mean (b$lifeExp)
```

# 6.7 select() subsets data column-wise (variables)

## We use select() to subset the data on variables or columns.

```{r}
select(gapminder, year, country, lifeExp) 
```

```{r}
select(gapminder, -continent, -lifeExp) # you can use - to deselect columns 
```

# 6.8 Use select() and filter() together

```{r}
gap_cambodia  <- filter(gapminder, country == "Cambodia")
```

```{r}
gap_cambodia2 <- select(gap_cambodia, -continent, -lifeExp)
```

# 6.9 Meet the new pipe %>% operator

## Before we go any further, we should explore the new pipe operator that dplyr imports from the magrittr package by Stefan Bache. This is going to change your life. You no longer need to enact multi-operation commands by nesting them inside each other. And we won’t need to make temporary variables like we did in the Cambodia example above. This new syntax leads to code that is much easier to write and to read: it actually tells the story of your analysis.

```{r}
gapminder %>% head()
```

```{r}
gapminder %>% head(3)
```

This means that instead of this:

## instead of this...
gap_cambodia  <- filter(gapminder, country == "Cambodia")
gap_cambodia2 <- select(gap_cambodia, -continent, -lifeExp) 

## ...we can do this
gap_cambodia  <- gapminder %>% filter(country == "Cambodia")
gap_cambodia2 <- gap_cambodia %>% select(-continent, -lifeExp) 

…But, we still have those temporary variables so we’re not truly that better off. But get ready to be majorly impressed:

# 6.9.1 Revel in the convenience

We can use the pipe to chain those two operations together:

```{r}
gap_cambodia  <- gapminder %>% 
  filter(country == "Cambodia") %>%
  select(-continent, -lifeExp) 
```

What’s happening here? In the second line, we were able to delete gap_cambodia2 <- gap_cambodia, and put the pipe operator above. This is possible since we wanted to operate on the gap_cambodia data anyways. And we weren’t truly excited about having a second variable named gap_cambodia2 anyways, so we can get rid of it. This is huge, because most of your data wrangling will have many more than 2 steps, and we don’t want a gap_cambodia14!

By using multiple lines I can actually read this like a story and there aren’t temporary variables that get super confusing. In my head:

    “start with the gapminder data, and then
    filter for Cambodia, and then
    drop the variables continent and lifeExp.”

Being able to read a story out of code like this is really game-changing. We’ll continue using this syntax as we learn the other dplyr verbs.

# 6.10 mutate() adds new variables

Imagine we want to know each country’s annual GDP. We can multiply pop by gdpPercap to create a new column named gdp.

```{r}
gapminder %>% mutate (gdp = pop * gdpPercap)
```

Calculate the population in thousands for all Asian countries in the year 2007 and add it as a new column.

```{r}
gapminder %>%
  filter(continent == "Asia",year == 2007) %>% mutate(pop_thousands_asia = pop/1000) %>% select(country, year, pop_thousands_asia) #this cleans up the dataframe but isn't necessary
```

# 6.11 group_by() operates on groups
<<<<<<< HEAD
=======

What if we wanted to know the total population on each continent in 2002? Answering this question requires a grouping variable.

```{r}
gapminder %>%
  filter(year == 2002) %>% group_by(continent) %>% mutate(cont_pop = sum(pop))
```

OK, this is great. But what if we don’t care about the other columns and we only want each continent and their population in 2002? Here’s the next function:

# 6.11.1 summarize() with group_by()

```{r}
gapminder %>%
  group_by(continent) %>%
  summarize(cont_pop = sum(pop)) %>%
  ungroup()
```

We can use more than one grouping variable. Let’s get total populations by continent and year.

```{r}
gapminder %>%
  group_by(continent, year) %>%
  summarize(cont_pop = sum(pop))
```

# 6.12 arrange() orders columns

This is ordered alphabetically, which is cool. But let’s say we wanted to order it in ascending order for year. The dplyr function is arrange().

```{r}
gapminder %>%
  group_by(continent, year) %>%
  summarize(cont_pop = sum(pop)) %>%
  arrange(year)
```

6.12.1 Your turn

    What is the maximum GDP per continent across all years?

```{r}
gapminder %>%
  mutate(gdp = pop * gdpPercap) %>%
  group_by(continent) %>%
  mutate(max_gdp = max(gdp)) %>%
  filter(gdp == max_gdp)
```


What if we wanted to know the total population on each continent in 2002? Answering this question requires a grouping variable.

```{r}
gapminder %>%
  filter(year == 2002) %>% group_by(continent) %>% mutate(cont_pop = sum(pop))
```

OK, this is great. But what if we don’t care about the other columns and we only want each continent and their population in 2002? Here’s the next function:

# 6.11.1 summarize() with group_by()

```{r}
gapminder %>%
  group_by(continent) %>%
  summarize(cont_pop = sum(pop)) %>%
  ungroup()
```

We can use more than one grouping variable. Let’s get total populations by continent and year.

```{r}
gapminder %>%
  group_by(continent, year) %>%
  summarize(cont_pop = sum(pop))
```

# 6.12 arrange() orders columns

This is ordered alphabetically, which is cool. But let’s say we wanted to order it in ascending order for year. The dplyr function is arrange().

```{r}
gapminder %>%
  group_by(continent, year) %>%
  summarize(cont_pop = sum(pop)) %>%
  arrange(year)
```

6.12.1 Your turn

    What is the maximum GDP per continent across all years?

6.12.1.1 Answer

```{r}
gapminder %>%
  mutate(gdp = pop * gdpPercap) %>%
  group_by(continent) %>%
  mutate(max_gdp = max(gdp)) %>%
  filter(gdp == max_gdp)
```

6.12.2 Your turn

1. arrange your data frame in descending order (opposite of what we’ve done). Expect that this is possible: ?arrange
2. save your data frame as a variable
3. find the maximum life expectancy for countries in Asia. What is the earliest year you encounter? The latest? Hint: you can use or base::max and dplyr::arrange()…
4. Knit your RMarkdown file, and sync it to GitHub (pull, stage, commit, push)


6.12.2.1 Answer 

```{r}
asia_life_exp <- gapminder %>%
  filter(continent == 'Asia') %>%
  group_by(country) %>%
  filter(lifeExp == max(lifeExp)) %>%
  arrange(year) ; asia_life_exp 
```

# my own version 1
```{r}
gapminder %>% filter (continent == "Asia") %>% select (lifeExp, country, continent, year) %>% group_by(country) %>% filter (lifeExp == max(lifeExp)) %>% arrange (lifeExp)
```

# my own version 2
```{r}
gapminder %>% filter (continent == "Asia") %>% select (lifeExp, country, continent, year) %>% group_by(country) %>% filter (lifeExp == max(lifeExp)) %>% arrange (desc(lifeExp))
```

# my own exercise 1
```{r}
gapminder %>% filter (country == "Indonesia") %>% select(country, year, gdpPercap, pop, lifeExp)
```

# my own exercise 2
```{r}
ina <- gapminder %>% filter (country == "Indonesia") %>% select(country, year, gdpPercap, pop, lifeExp) ; ggplot(data = ina) +
    geom_point(aes(x = year, y = gdpPercap, color = pop))
```

# my own exercise 3
```{r}
ggplot(data = gapminder) +
    geom_point(aes(x = pop, y = gdpPercap))
```

# 6.13 All together now

```{r}
gap_max_gdp <- gapminder %>% 
  select(-continent, -lifeExp) %>% # or select(country, year, pop, gdpPercap)
  group_by(country) %>%
  mutate(gdp = pop * gdpPercap) %>%
  summarize(max_gdp = max(gdp)) ; gap_max_gdp
```

# my own version 1

```{r}
gap_max_gdp <- gapminder %>% 
  group_by(country) %>%
  mutate(gdp = pop * gdpPercap) %>%
  summarize(max_gdp = max(gdp)) ; gap_max_gdp
```

# my own version 2

```{r}
gap_max_gdp_cont <- gapminder %>% 
  group_by(continent, country, year) %>%
  mutate(gdp_cont = pop * gdpPercap) %>%
  summarize(max_gdp_cont = max(gdp_cont)) %>% arrange (desc(max_gdp_cont)); gap_max_gdp_cont
```
# my own version 3

```{r}
gap_max_gdp_cont <- gapminder %>% 
  group_by(continent) %>%
  mutate(gdp_cont = pop * gdpPercap) %>%
  summarize(max_gdp_cont = max(gdp_cont)) %>% arrange (desc(max_gdp_cont)); gap_max_gdp_cont
```

# 6.14 Joining datasets

## read in the data. (same URL as yesterday, with co2.csv instead of gapminder.csv)

```{r}
co2 <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/co2.csv")
```

```{r}
## explore
co2 %>% head()
co2 %>% dim() # 12
```

```{r}
## create new variable that is only 2007 data
gap_2007 <- gapminder %>%
  filter(year == 2007) 
gap_2007 %>% dim() # 142  
```

```{r}
## left_join gap_2007 to co2
lj <- left_join(gap_2007, co2, by = "country")
```

```{r}
## explore
lj %>% dim() #142
lj %>% summary() # lots of NAs in the co2_2007 columm
lj %>% View()
```

```{r}
## right_join gap_2007 and co2
rj <- right_join(gap_2007, co2, by = "country")
```


```{r}
## explore
rj %>% dim() # 12
rj %>% summary()
rj %>% View() 
```

